---
title: "Visualizing Plastics: Waste Composition and Environmental Cost"
author: "Nathan Lee, Noah Nayve, Ryan Ekdahl, Taiga Sobajima"
format:
  dashboard:
    scrolling: true
    orientation: rows
    theme: flatly
    logo: "logo.png"
    nav-buttons:
      - icon: envelope
        href: mailto:nlee174@calpoly.edu
        aria-label: Nathan Lee
      - icon: envelope
        href: mailto:nnayve@calpoly.edu
        aria-label: Noah Nayve
      - icon: envelope
        href: mailto:rekdahl@calpoly.edu
        aria-label: Ryan Ekdahl
      - icon: envelope
        href: mailto:tsobajim@calpoly.edu
        aria-label: Taiga Sobajima
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(glue)
library(gt)
library(plotly)
library(scales)
library(ggridges)
library(broom)
library(ggrepel)
library(rvest)
library(httr)
library(geodata)
library(leaflet)
library(sf)
```

```{r}
#| label: load-data
#| message: false

population_plastics <- read_csv("data/population_plastics.csv") |>
  mutate(country = case_when(
    country == "United States Of America" ~ "United States of America",
    country == "Taiwan_ Republic Of China (Roc)" ~ "Taiwan",
    TRUE ~ country
  ))
```

```{r}
#| label: prep
#| message: false
#| warning: false

summarize_plastic_and_ecology <- function(country_name, data = population_plastics) {
  if (!(country_name %in% data$country)) stop(paste("Country not found:", country_name))
  country_stats <- data |>
    filter(country == country_name, parent_company == "Grand Total", grand_total > 0) |>
    summarize(
      country = country_name,
      avg_pet = mean(pet / grand_total, na.rm = TRUE),
      avg_hdpe = mean(hdpe / grand_total, na.rm = TRUE),
      avg_ldpe = mean(ldpe / grand_total, na.rm = TRUE),
      avg_pp = mean(pp / grand_total, na.rm = TRUE),
      avg_ps = mean(ps / grand_total, na.rm = TRUE),
      avg_pvc = mean(pvc / grand_total, na.rm = TRUE),
      avg_o = mean(o / grand_total, na.rm = TRUE),
      avg_forested = mean(forested_area, na.rm = TRUE),
      avg_threatened = mean(threatened_species, na.rm = TRUE)
    )
  leading_plastics <- country_stats |>
    pivot_longer(cols = !c(country, avg_forested, avg_threatened),
                 names_to = "plastic_type", values_to = "avg_value") |>
    slice_max(order_by = avg_value, n = 1, with_ties = FALSE) |>
    mutate(plastic_type = str_remove(plastic_type, "avg_"),
           plastic_type = str_to_upper(plastic_type),
           plastic_type = if_else(plastic_type == "O", "Other", plastic_type)) |>
    pull(plastic_type)
  country_stats |> mutate(leading_plastic = leading_plastics)
}

# precompute all needed data
all_countries_rq2 <- population_plastics |>
  filter(parent_company == "Grand Total", grand_total > 0,
         !is.na(forested_area), !is.na(threatened_species)) |>
  distinct(country) |> pull(country)

plastic_ecology_summary_table <- map(all_countries_rq2, summarize_plastic_and_ecology) |>
  bind_rows()

top3_plastics <- population_plastics |>
  filter(parent_company == "Grand Total") |>
  summarize(across(c(pet, hdpe, ldpe, pp, ps, pvc), ~ sum(.x, na.rm = TRUE))) |>
  pivot_longer(everything(), names_to = "plastic", values_to = "total") |>
  slice_max(total, n = 3) |> pull(plastic)

gt_style <- function(gt_obj) {
  gt_obj |>
    tab_style(style = cell_fill(color = "#457B9D"), locations = cells_column_labels()) |>
    tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_labels()) |>
    tab_style(style = cell_fill(color = "#1D3557"), locations = cells_column_spanners()) |>
    tab_style(style = cell_text(color = "white", weight = "bold"), locations = cells_column_spanners())
}
```

# Main

## Row 1 {height="500px"}

### Bubble Plot {width="70%" .tabset}

#### Overview
f

#### Bubble Plots
bubble plot will be here
x axis = plastic type count
y axis = 

### General {width="30%"}

#### Value Boxes

```{r}
#| content: valuebox
#| title: "Countries Analyzed"

list(
  icon = "globe",
  color = "#D3D3D3",
  value = n_distinct(all_countries_rq2)
)
```

```{r}
#| content: valuebox
#| title: "Most Common Plastic Collected"

list(
  icon  = "recycle",
  color = "#D3D3D3",
  value = "PET"
)
```

```{r}
#| content: valuebox
#| title: "Significant Predictor of PET"

list(
  icon  = "bug",
  color = "#D3D3D3",
  value = "Species"
)
```

## Row 2

### Density Ridge {width="50%"}

::: {.card}
```{r}
#| message: false
#| label: ridge plot for threatened species
#| fig-alt: "Ridge plot showing threatened species distribution across top 3 plastic types"

population_plastics |>
  filter(parent_company == "Grand Total",
         grand_total > 0,
         !is.na(threatened_species)
         ) |>
  pivot_longer(cols = all_of(top3_plastics),
               names_to = "plastic_type",
               values_to = "count") |>
  filter(count > 0) |>
  mutate(plastic_type = factor(toupper(plastic_type),
                               levels = toupper(top3_plastics))
         ) |>
  ggplot(aes(x = threatened_species,
             y = plastic_type,
             fill = plastic_type,
             weight = count)
         ) +
  geom_density_ridges(scale = 0.8, alpha = 0.7) +
  scale_x_log10(labels = scales::label_comma()) +
  scale_fill_manual(values = c("#E8A0A0", "#C45C5C", "#8B1A1A")) +
  labs(title = "Threatened Species Counts Across Top 3 Plastic Types",
       subtitle = "Each country weighted by quantity of plastic collected",
       x = "Threatened Species Count (log scale)", y = "") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(size = 11),
        legend.position = "none")
```
PP is collected in the most biodiverse countries, while PET dominates in nations with fewer threatened species. LDPE falls in between.
:::

### Bar Charts {width="50%"}

::: {.panel-tabset}
#### Low Coverage

```{r}
#| fig-alt: "Bar chart showing bottom 5 forest coverage countries with plastic waste breakdown"
#| label: bottom 5 countries bar chart with plastic waste breakdown

bottom5 <- plastic_ecology_summary_table |>
  mutate(across(everything(),
                ~replace_na(., 0))
         ) |>
  filter(!is.na(avg_forested)) |>
  slice_min(avg_forested, n = 5)

bottom5_forest <- ggplot(bottom5, aes(x = reorder(country, avg_forested),
                                      y = avg_forested,
                                      text = glue("PET: {round(avg_pet * 100, 2)}%
                                      HDPE: {round(avg_hdpe * 100, 2)}%
                                      LDPE: {round(avg_ldpe * 100, 2)}%
                                      PP: {round(avg_pp * 100, 2)}%
                                      PS: {round(avg_ps * 100, 2)}%
                                      PVC: {round(avg_pvc * 100, 2)}%
                                      Other: {round(avg_o * 100, 2)}%"))) +
  geom_bar(stat = "identity",
           fill = "#F3D4D5") +
  coord_flip() +
  labs(y = "Forest Coverage (%)",
       x = "",
       title = "Bottom 5 Forest Coverage Countries") +
  theme_minimal()

ggplotly(bottom5_forest, tooltip = "text")
```

#### Medium Coverage

```{r}
#| fig-alt: "Bar chart showing medium 5 forest coverage countries with plastic waste breakdown"
#| label: medium 5 countries bar chart with plastic waste breakdown

medium5 <- plastic_ecology_summary_table |>
  mutate(across(everything(),
                ~replace_na(., 0))
         ) |>
  filter(!is.na(avg_forested),
         avg_forested >= 19) |>
  slice_min(avg_forested, n = 5)

medium5_forest <- ggplot(medium5, aes(x = reorder(country, avg_forested),
                                      y = avg_forested,
                                      text = glue("PET: {round(avg_pet * 100, 2)}%
                                      HDPE: {round(avg_hdpe * 100, 2)}%
                                      LDPE: {round(avg_ldpe * 100, 2)}%
                                      PP: {round(avg_pp * 100, 2)}%
                                      PS: {round(avg_ps * 100, 2)}%
                                      PVC: {round(avg_pvc * 100, 2)}%
                                      Other: {round(avg_o * 100, 2)}%"))) +
  geom_bar(stat = "identity",
           fill = "#F3D4D5") +
  coord_flip() +
  labs(y = "Forest Coverage (%)",
       x = "",
       title = "Medium 5 Forest Coverage Countries") +
  theme_minimal()

ggplotly(medium5_forest, tooltip = "text")
```

#### High Coverage

```{r}
#| fig-alt: "Bar chart showing top 5 forest coverage countries with plastic waste breakdown"
#| label: top 5 countries bar chart with plastic waste breakdown

high5 <- plastic_ecology_summary_table |>
  mutate(across(everything(),
                ~replace_na(., 0))
         ) |>
  filter(!is.na(avg_forested),
         avg_forested >= 36) |>
  slice_max(avg_forested, n = 5)

high5_forest <- ggplot(high5, aes(x = reorder(country, avg_forested),
                                      y = avg_forested,
                                      text = glue("PET: {round(avg_pet * 100, 2)}%
                                      HDPE: {round(avg_hdpe * 100, 2)}%
                                      LDPE: {round(avg_ldpe * 100, 2)}%
                                      PP: {round(avg_pp * 100, 2)}%
                                      PS: {round(avg_ps * 100, 2)}%
                                      PVC: {round(avg_pvc * 100, 2)}%
                                      Other: {round(avg_o * 100, 2)}%"))) +
  geom_bar(stat = "identity",
           fill = "#F3D4D5") +
  coord_flip() +
  labs(y = "Forest Coverage (%)",
       x = "",
       title = "High 5 Forest Coverage Countries") +
  theme_minimal()

ggplotly(high5_forest, tooltip = "text")
```
:::

# India

```{r}
#| label: getting India data

URL <- "https://www.pib.gov.in/PressReleasePage.aspx?PRID=1943210&reg=3&lang=2"

# used AI here to help get url, basically disguises so we can fetch the data
page <- GET(URL, user_agent("Mozilla/5.0 (Windows NT 10.0; Win64; x64)"))

tables_india <- content(page, "text") |>
  read_html() |>
  html_table(fill = TRUE)

plastic_raw <- tables_india[[2]]

plastic_india <- plastic_raw |>
  select(state = 2, waste_2017_18 = 4, waste_2018_19 = 5) |>
  slice(-1, -2) |>
  mutate(waste_2017_18 = as.numeric(waste_2017_18),
         waste_2018_19 = as.numeric(waste_2018_19)) |>
  mutate(state = recode(state,
    "Jammu & Kashmir"                          = "Jammu and Kashmir",
    "Madhya  Pradesh"                          = "Madhya Pradesh",
    "Andaman and Nicobar Islands"              = "Andaman & Nicobar",
    "Dadra and Nagar Haveli and Daman and Diu" = "Dadra and Nagar Haveli",
    "Delhi NCR"                                = "NCT of Delhi"
  ))

if (!file.exists("data/india_gadm.rds")) {
  india_GADM <- gadm("IND", level = 1, path = "data/") |> st_as_sf()
  saveRDS(india_GADM, "data/india_gadm.rds")
} else {
  india_GADM <- readRDS("data/india_gadm.rds")
}

india_df <- india_GADM |>
  left_join(plastic_india, by = c("NAME_1" = "state"))
```

## Row 1 {height="30%"}

```{r}
#| content: valuebox
#| label: India is country chosen
#| title: "Country Chosen"

list(
  icon = "geo-fill",
  color = "#D3D3D3",
  value = "India"
)
```

```{r}
#| content: valuebox
#| label: number of India threatened species
#| title: "Threatened Species Count"

list(
  icon = "exclamation-triangle",
  color = "#D3D3D3",
  value = "1,118"
)
```

```{r}
#| content: valuebox
#| label: India forested area
#| title: "Average Forested Area"

india_forest <- plastic_ecology_summary_table |>
  filter(country == "India") |>
  pull(avg_forested)

list(
  icon = "tree",
  color = "#D3D3D3",
  value = paste0(india_forest, "%")
)
```

## Row 2 {height="70%"}

### Column 1

::: {.card}
India has a high forested area and threatened species count. Also, they have the most complete and distributed plastic profile in the data.

```{r}
#| label: bar chart for India plastic breakdown
#| fig-height: 7
#| fig-width: 15

just_india <- plastic_ecology_summary_table |>
  filter(country == "India") |>
  select(-country, -avg_forested, -avg_threatened, -leading_plastic) |>
  pivot_longer(cols = everything(),
               names_to = "plastic_type",
               values_to = "proportion") |>
  mutate(plastic_type = str_remove(plastic_type, "avg_") |> str_to_upper()) |>
  mutate(plastic_type = case_when(plastic_type == "O" ~ "Other",
                                  TRUE ~ plastic_type))

india_colors <- c("PVC"  = "#880e4f",
                  "PS" = "#ad1457",
                  "PP" = "#c2185b",
                  "PET"   = "#e91e63",
                  "Other"   = "#f48fb1",
                  "LDPE" = "#ffb3c6",
                  "HDPE"  = "#fce4ec"
                  )

ggplot(just_india, aes(x = proportion,
                          y = plastic_type,
                          fill = plastic_type)) +
  geom_col() +
  geom_text(aes(label = paste0(round(proportion * 100, 1), "%")),
            hjust = -0.2,
            size  = 4) +
  scale_fill_manual(values = india_colors) +
  labs(title = "Plastic Type Distribution — India",
       x     = "Proportion",
       y     = ""
       ) +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.y = element_text(size = 18),
        axis.text.x = element_text(size = 15),
        axis.title.x = element_text(size = 18),
        plot.title = element_text(size = 25, face = "bold"))
```
:::

### Column 2

```{r}
#| label: map with states and territories

pal <- colorNumeric("RdPu", domain = india_df$waste_2018_19, na.color = "#cccccc")

leaflet(india_df) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  setView(lng = 90, lat = 22, zoom = 3.5) |>
  addPolygons(fillColor = ~pal(waste_2018_19),
              fillOpacity = 0.8,
              color = "black",
              weight = 1.5,
              label = ~paste0(NAME_1, ": ",
                              ifelse(is.na(waste_2018_19), "No data",
                                     paste(formatC(waste_2018_19, format="f", digits=0, big.mark=","),
                                           "tonnes"))),
              highlightOptions = highlightOptions(weight = 2, fillOpacity = 0.9)
              ) |>
  addLegend(pal = pal,
            values = ~waste_2018_19,
            title    = "Plastic waste (tonnes, 2018-19)",
            position = "bottomright",
            na.label = "No data")
```
